---
title: "Panel regression with R"
subtitle: "Panel Data Analysis, ECPR Summer School 2018"
author: "Akos Mate"
date: '2018 July'
output:
    html_document:
        toc: true
        toc_depth: 3
        theme: readable
        css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      collapse = TRUE,
                      comment = "#>",
                      message = FALSE
)
```


> Resources to check if you are interested:  
> * [Princeton Panel 101 with R](https://www.princeton.edu/~otorres/Panel101R.pdf)    
> * [`plm` package walkthrough](https://cran.r-project.org/web/packages/plm/vignettes/plm.pdf)    
> * [MIT econometrics companion: 'Econometrics in R'](https://ocw.mit.edu/courses/economics/14-381-statistical-method-in-economics-fall-2013/study-materials/MIT14_381F13_EcnomtrisInR.pdf)

This is a companion code to the Stata seminar sessions for the Panel Regression course at ECPR Methods Summer School 2018, CEU. We will not go into details of R, just focus on how to run panel regressions in R. If you are interested in R in general, you can check the [Github repository](https://github.com/aakosm/r_basics_ecpr18) for the 'R Basics' short course.


```{r}
library(dplyr) # data manipulation
library(haven) # reading .dta files
library(broom) # handling regression output
library(plm) # panel regression package
library(pcse) # panel corrected standard error

```

First, I checked if the R functions and the Stata `xtreg` give similar results by re-running Princeton's Stata panel 101 course, found here: [https://www.princeton.edu/~otorres/Panel101.pdf](https://www.princeton.edu/~otorres/Panel101.pdf)

The data:
```{r}
panel_data <- read_stata("http://dss.princeton.edu/training/Panel101.dta")

panel_data <- pdata.frame(panel_data, index = c("country", "year"))

glimpse(panel_data)

pdim(panel_data)
```

Two ways to set up your panel data with the `plm` package. Either add an in function specificiation with the `index = c("individual", "time")` or transform your data with the `pdata.frame` function.

# 1. lm()/plm() vs xtreg

## 1.1 OLS

`lm(y ~ x1, data = panel_data)` = `regress y x1`

```{r}
summary(lm(y ~ x1, data = panel_data))
```



Using the `plm` package to do pooled OLS.
`plm(y ~ x1, data = panel_data, model = "pooling")` = `regress y x1`

```{r}
summary(plm(y ~ x1, data = panel_data, model = "pooling"))
```


Using country dummies in our OLS. In R, adding a `factor(x)` makes the `lm()` function take it as a dummy.

```{r}
summary(lm(y ~ x1 + factor(country), data = panel_data))
```


`lm(y ~ x1 + factor(country), data = panel_data)` =

* `. xi: regress y x1 i.country`

* `. xtreg y x1, fe`

* `. areg y x1, absorb(country)`


## 1.2 Fixed and mixed effects models

We can estimate fixed effects regressions by specifying the `model = ` in the `plm()` function.

```{r}
summary(plm(y ~ x1, data = panel_data, model = "within"))
```



`plm(y ~ x1, data = panel_data, model = "within")` =

* `. xi: regress y x1 i.country`,

* `. xtreg y x1, fe`,

* `. areg y x1, absorb(country)`



```{r}
summary(plm(y ~ x1, data = panel_data, model = "random"))
```


`plm(y ~ x1, data = panel_data, model = "random")` = `. xtreg y x1, re`


## 1.3 Hausman test

```{r}
fixed <- plm(y ~ x1, data = panel_data, model = "within")
random <- plm(y ~ x1, data = panel_data, model = "random")

phtest(fixed, random)
```

`phtest` = `.hausman`


# 2. Replication data

We will use the .dta file and check if the imported data is in the data frame format, that the `plm` package requres to run panel models.

```{r}
ldc_data <- read_stata("LDC_IO_replication.dta")
```

Let's check what we've loaded

```{r}
glimpse(ldc_data)

is.data.frame(ldc_data)

pdim(ldc_data)
```

## 2.1 Pooled OLS

Looks great, now we can dive into `plm`. An exhaustive write-up of the package's use is here: [https://cran.r-project.org/web/packages/plm/vignettes/plm.pdf](https://cran.r-project.org/web/packages/plm/vignettes/plm.pdf). In R, we set the indexing within the `plm()` function, as opposed to Stata's `xtset`.


```{r}
options(scipen = 999)


ols1 <- lm(newtar ~ l1polity + l1lnpop + l1gdp_pc + date, data = ldc_data)
summary(ols1)


ols2 <- lm(newtar ~ l1polity + l1lnpop + l1gdp_pc + date + factor(country), data = ldc_data)
summary(ols2)
```

### 2.1.1 Panel corrected standard errors

To get the panel corrected standard errors, we should use the `pcse` package. We use the `ols1` result as our input for the `pcse()` function.

```{r, eval=FALSE}
ols_pcse <- pcse(ols1, groupN = ldc_data$country, groupT = ldc_data$date)

summary(ols_pcse)
```


## 2.2 Fixed effects and first differencing

### 2.2.1 Fixed effects

You can specify the effects, after the `effect = `. Possible values:
* `"individual"` (default for FE)
* `"time"`
* `"twoways"`
* `"nested"`

```{r}
fe <- plm(newtar ~ l1polity + date + l1lnpop, data = ldc_data, model = "within", index = c("date", "country"), effect = "twoway")

summary(fe)
```

Or using the `lm` function and adding country dummies we ge the same results as the `xtreg newtar l1polity l1lnpop l1gdp_pc date, fe` in our Stata do file. A slight difference in the intercept because of all the estimated country dummies.

```{r}
summary(lm(newtar ~ l1polity + l1lnpop + l1gdp_pc + date + factor(country), data = ldc_data))
```


### 2.2.2 First differencing

```{r}
fd <- plm(newtar ~ l1polity + l1lnpop + l1gdp_pc, data = ldc_data, model = "fd")

summary(fd)
```

### 2.2.3 Clustered standard errors

Let's use our fe model to get cluster robust standard errors.

```{r}
lmtest::coeftest(fe, vcov=vcovHC(fe,type="HC0",cluster="group"))
```


## 2.3 Random effects

You can specify the method of estimation for the variance components in the RE model:
* `"swar"` (default) Swamy and Arora (1972)
* `"ammiya"` Amemiya (1971)
* `"walhus"` Wallace and Hussain (1969)
* `"nerlove"` Nerlove (1971)

```{r, error=TRUE}
re1 <- plm(newtar ~ l1polity + l1lnpop + l1gdp_pc + date, data = ldc_data, model = "random", random.method = "swar")


```
Use gls (generalized least squares) estimation (similarly to `xtreg re` in Stata)

```{r}
re2 <- pggls(newtar ~ l1polity + l1lnpop + l1gdp_pc + date, data = ldc_data, model = "pooling")

summary(re2)

```

## 2.4 IV estimation


`plm` can deal with instruments as well, specified in the formula argument after `|` sign or using the `instruments = ` argument.

```{r, error=TRUE}
iv <- plm(newtar ~ l1polity + l1lnpop + l1gdp_pc + l1partyage2000 + l1ssch, instruments = ~l1partyage2000 + l1ssch,  data = ldc_data, model = "random")

iv2 <- plm(newtar ~ l1polity + l1lnpop + l1gdp_pc + l1partyage2000 + l1ssch | + l1partyage2000 + l1ssch,  data = ldc_data, model = "random")
```


# Hausman-test

```{r, eval=FALSE}
phtest(fe1, re1)
```
